---
title: "Implementation of Shi's non-degeranate Vuong test"
author: Yves Croissant
header-includes:
  - \usepackage{xcolor}
output: 
  rmarkdown::html_vignette:
toc: false
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Implementation of Shi's non-degeranate Vuong test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# The original Vuong test

@VUON:89 proposed a test for non-nested model. He considers two
competing models, $F_\beta = \left\{f(y|z; \beta); \beta \in
\beta\right\}$ and $G_\gamma = \left\{g(y|z; \gamma); \gamma \in
\Gamma\right\}$. Denoting $h(y | z)$ the true conditional density.
The distance of $F_\beta$ from the true model is measured by the
minimum KLIC:

$$
D_f = \mbox{E}^0\left[\ln h(y\mid z)\right] - \mbox{E}^0\left[\ln f(y\mid z;
\beta_*)\right]
$$

where $\mbox{E}^0$ is the expected value using the true joint
distribution of $(y, X)$ and $\beta_*$ is the pseudo-true value of
$\beta$^[$\beta_*$ is called the pseudo-true value because $f$ may
be an uncorrect model.]. As the true model is unobserved, denoting
$\theta^\top = (\beta ^ \top, \gamma ^ \top)$, we consider the
difference of the KLIC distance to the true model of model $G_\gamma$
and model $F_\beta$:


$$ \Lambda(\theta) = D_g - D_f = \mbox{E}^0\left[\ln f(y\mid z;
\beta_*)\right]- \mbox{E}^0\left[\ln g(y\mid z; \gamma_*)\right] =
\mbox{E}^0\left[\ln \frac{f(y\mid z; \beta_*)}{g(y\mid z;
\gamma_*)}\right] $$

The null hypothesis is that the distance of the two models to the true
models are equal or, equivalently, that: $\Lambda=0$. The alternative
hypothesis is either $\Lambda>0$, which means that $F_\beta$ is
better than $G_\gamma$ or $\Lambda<0$, which means that $G_\gamma$
is better than $F_\beta$. Denoting, for a given random sample of
size $N$, $\hat{\beta}$ and $\hat{\gamma}$ the maximum likelihood
estimators of the two models and $\ln L_f(\hat{\beta})$ and $\ln
L_g(\hat{\gamma})$ the maximum value of the log-likelihood functions
of respectively $F_\beta$ and $G\gamma$, $\Lambda$ can be
consistently estimated by:

$$
\hat{\Lambda}_N = \frac{1}{N} \sum_{n = 1} ^ N  \left(\ln f(y_n \mid
x_n, \hat{\beta}) - \ln g(y_n \mid x_n, \hat{\gamma})\right)
= \frac{1}{N} \left(\ln L_f(\hat{\beta}) - \ln L_g(\hat{\gamma})\right)
$$

which is the likelihood ratio divided by the sample size. Note that
the statistic of the standard likelihood ratio test, suitable for
nested models is $2 \left(\ln L^f(\hat{\beta}) - \ln
L^g(\hat{\gamma})\right)$, which is $2 N \hat{\Lambda}_N$.

The variance of $\Lambda$ is:

$$
\omega^2_* = \mbox{V}^o \left[\ln \frac{f(y \mid x; \beta_*)}{g(y
\mid x; \gamma_*)}\right]
$$

which can be consistently estimated by:

$$
\hat{\omega}_N^2 = \frac{1}{N} \sum_{n = 1} ^ N  \left(\ln f(y_n \mid
x_n, \hat{\beta}) - \ln g(y_n \mid x,_n \hat{\gamma})\right) ^ 2 - 
\hat{\Lambda}_N ^ 2
$$

Three different cases should be considered:

- when the two models are nested, $\omega^2_*$ is necessarily 0,
- when the two models are overlapping (which means than the models
  coincides for some values of the parameters), $\omega^2_*$ *may be*
  equal to 0 or not,
- when the two models are stricly non-nested, $\omega^2_*$ is
  necessarely strictly positive.

The distribution of the statistic depends on whether $\omega^2_*$ is
zero or positive.  If $\omega^2_*$ is positive, the statistic is
$\hat{T}_N = \sqrt{N}\frac{\hat{\Lambda}_N}{\hat{\omega}_N}$ and,
under the null hypothesis that the two models are equivalent,
follows a standard normal distribution. This is the case for two
strictly non-nested models.

On the contrary, if $\omega^2_* = 0$, the distribution is much more
complicated. We need to define two matrices: $A$ contains the expected
values of the second derivates of $\Lambda$:

$$
A(\theta_*) = \mbox{E}^0\left[\frac{\partial^2 \Lambda}{\partial \theta
\partial \theta ^ \top}\right] = 
\mbox{E}^0\left[\begin{array}{cc}
\frac{\partial^2 \ln f}{\partial \beta \partial \beta ^
\top} & 0 \\
0 & -\frac{\partial^2 \ln g}{\partial \beta \partial \beta ^
\top}
\end{array}\right]
=
\left[
\begin{array}{cc}
A_f(\beta_*) & 0 \\
0 & - A_g(\gamma_*)
\end{array}
\right]
$$

and $B$ the variance of its first derivatives:

$$
B(\theta_*) =
\mbox{E}^0\left[\frac{\partial \Lambda}{\partial
\theta}\frac{\partial \Lambda}{\partial \theta ^ \top}\right]=
\mbox{E}^0\left[
\left(\frac{\partial \ln f}{\partial \beta}, 
- \frac{\partial \ln g}{\partial \gamma} \right)
\left(\frac{\partial \ln f}{\partial \beta ^ \top}, 
- \frac{\partial \ln g}{\partial \gamma ^ \top} \right)
\right]
= \mbox{E}^0\left[
\begin{array}{cc}
\frac{\partial \ln f}{\partial \beta} \frac{\partial \ln f}{\partial
\beta^\top} & 
- \frac{\partial \ln f}{\partial \beta} \frac{\partial \ln g}{\partial
\gamma ^ \top} \\
- \frac{\partial \ln g}{\partial \gamma} \frac{\partial \ln f}{\partial
  \beta^\top} & 
\frac{\partial \ln g}{\partial \gamma} \frac{\partial \ln g}{\partial \gamma^\top}
\end{array}
\right]
$$

or:

$$
B(\theta_*) =  
\left[
\begin{array}{cc}
B_f(\beta_*) & - B_{fg}(\beta_*, \gamma_*) \\
- B_{gf}(\beta_*, \gamma_*) & B_g(\gamma_*)
\end{array}
\right]
$$

Then:

$$
W(\theta_*) =  B(\theta_*) \left[-A(\theta_*)\right] ^ {-1}= 
\left[
\begin{array}{cc}
-B_f(\beta_*) A^{-1}_f(\beta_*) & - B_{fg}(\beta_*, \gamma_*)
A^{-1}_g(\gamma_*) \\
B_{gf}(\gamma_*, \beta_*) A^{-1}_f(\beta_*) & B_g(\gamma_*)
A^{-1}_g(\gamma_*)
\end{array}
\right]
$$

Denote $\lambda_*$ the eigen values of $W$.  When $\omega_*^2 = 0$
(which is always the case for nested models), the statistic is the one
used in the standard likelihood ratio test: $2 (\ln L_f - \ln L_g) = 2
N \hat{\Lambda}_N$ which, under the null, follows a weighted $\chi ^
2$ distribution with weights equal to $\lambda_*$. The Vuong test can
be seen in this context as a more robust version of the standard
likelihood ratio test, because it doesn't assume, under the null, that
the larger model is correctly specified.

Note that, if the larger model is correctly specified, the information
matrix equality implies that $B_f(\theta_*)-A_f(\theta_*)$. In this
case, the two matrices on the diagonal of $W$ reduce to $-I_{K_f}$ and
$I_{K_g}$, the trace of $W$ to $K_g - K_f$ and the distribution of the
statistic under the null reduce to a $\chi^2$ with $K_g - K_f$ degrees
of freedom.

The $W$ matrice can be consistently estimate by computing the first
and the second derivatives of the likelihood functions of the two
models for $\hat{\theta}$. For example, 

$$
\hat{A}_f(\hat{\beta}) = \frac{1}{N}
\sum_{n= 1} ^ N \frac{\partial^2 \ln f}{\partial \beta \partial
\beta ^ \top}(\hat{\beta}, x_n, y_n)
$$

$$ \hat{B}_{fg}(\hat{\theta})= \frac{1}{N} \sum_{n=1}^N
\frac{\partial \ln f}{\partial \beta}(\hat{\beta}, x_n, y_n)
\frac{\partial \ln g}{\partial \gamma^\top}(\hat{\gamma}, x_n, y_n)
$$

For the overlapping case, the test should be performed in two steps:

- the first step consists on testing whether $\omega_*^*$ is 0 or
  not. This hypothesis is based on the statistic $N \hat{\omega} ^ 2$
  which, under the null ($\omega_*^2=0$) follows a weighted $\chi ^ 2$
  distributions with weights equal to $\lambda_* ^ 2$. If the null
  hypothesis is not rejected, the test stops at this step and the
  conclusion is that the two models are equivalent,
- if the null hypothesis is reject, the second step consists on
  applying the test for non-nested models previously described.


# The non-degenerate Vuong test

@SHI:15 proposed a non-degenerate version of the @VUON:89 test. She
shows that the Vuong test has size distortion, leading to subsequent
overrejection. The cause of this problem is that the distribution of
$\hat{\Lambda}$ is discontinuous in the $\omega^2$ parameter (namely a
normal distribution if $\omega^2 > 0$ and a distribution related to a
weight $\chi^2$ distribution if $\omega^2 = 0$). Especially in small
samples, it may be difficult to distinguish a positive versus a zero
value of $\omega ^ 2$ because of sampling error. To solve this
problem, using local asymptotic theory, @SHI:15 showed that, rewriting
the Vuong statistic as:

$$
\hat{T} = \frac{N \hat{\Lambda}_N}{\sqrt{N \hat{\omega} ^ 2_N}}
$$

the asymptotic distribution of the numerator and of the square of the
denominator of the Vuong statistic is the same as:

$$
\left(
\begin{array}{cc} 
N \hat{\Lambda}_N \\ N \hat{\omega} ^ 2 _ N
\end{array}
\right)
\rightarrow^d 
\left(
\begin{array}{cc} 
J_\Lambda \\ J_\omega 
\end{array}
\right) 
= 
\left(
\begin{array}{cc} 
\sigma z_\omega - z_\theta ^ \top V z_\theta / 2 \\ 
\sigma ^ 2 - 2 \sigma \rho_* ^ \top V z_\theta + z_\theta ^ \top V ^ 2
z_\theta
\end{array}
\right)
$$

where:

$$
\left(\begin{array}{c}z_\omega \\ z_\theta \end{array}\right) \sim
N \left(0, \left(\begin{array}{cc} 1 & \rho_* ^ \top \\ \rho_* & I \end{array}\right) \right),
$$

$\rho_*$ is a vector of length $K_f + K_g$, $\sigma$ a positive scalar
and V is the diagonal matrix containing the eigen values of $B ^
{\frac{1}{2}} A ^ {-1} B ^ {\frac{1}{2}}$.


Based on this result, @SHI:15 showed:

- that the expected value of the numerator is $-\mbox{trace}(V) / 2$,
  the classical Vuong statistic is therefore biased and this bias can
  be severe in small samples and when the degree of parametrization of
  the two models are very different^[As the trace of V is the same as
  the trace of $A ^ {-1} B$, when the information matrix identity
  holds, it is equal to $-K_f + K_g$. The bias of the numerator is
  therefore caused by the difference in the degree of parametrization
  of the two models.],
- that the denominator, being random, can take values close to zero
  with a significant probability, which can generate fat tails in the
  distribution of the statistic.


@SHI:15 therefore proposed
to modify the numerator of the Vuong statistic:

$$\hat{\Lambda}^{\mbox{mod}}_N = \hat{\Lambda}_N + \frac{\mbox{tr}(V)}{2 N}$$

and to add a constant to the denominator, so that:

$$
\left(\hat{\omega}^{\mbox{mod}}(c)\right) ^ 2 = \hat{\omega} ^ 2 + c \;
\mbox{tr}(V) ^ 2 / N
$$

The non-degenarate Vuong test is then:

$$
T_N^{\mbox{mod}} =
\frac{\hat{\Lambda}^{\mbox{mod}}_N}{\hat{\omega}^{\mbox{mod}}}=
\sqrt{N}\frac{\hat{\Lambda}_N + \mbox{tr}(V) / 2N}{\sqrt{\hat{\omega}
^ 2 + c \;\mbox{tr}(V) ^ 2 / N}}
$$

The distribution of the modified Vuong statistic can be estimated by
simulations: drawing in the distribution of $(z_\omega,
z_\theta^\top)$, we compute for every draw $J_\Lambda$, $J_\omega$ and
$J_\Lambda / \sqrt{J_\omega}$.  As $\sigma$ and $\rho_*$ can't be
estimated consistently, the supremum other these parameters are taken,
and @SHI:15 indicates that $\rho_*$ should be in this case a vector
where all the elements are zero except for the one that coincides with
the highest absolute value of $V$ which is set to 1.

The Shi test is then computed as follow:

@. start with a given size for the test, say $\alpha = 0.05$,
@. for a given value of $c$, choose $\sigma$ which maximize the
simulated critical value for $c$ and $\alpha$,
@. adjust $c$ so that this critical value equals the normal critical
value, up to a small disperency (say 0.1); for example, if the size
is 5%, the target is $v_{1 - \alpha / 2} = 1.96 + 0.1 = 2.06$,
@. compute $\hat{T}_N^{\mbox{mod}}$ for the given values of $c$ and $\sigma$
; if $\hat{T}_N^{\mbox{mod}} > v_{1 - \alpha / 2}$, reject the null
hypothesis at the $\alpha$ level,
@. to get a p-value, if $\hat{T}_N^{\mbox{mod}} > v_{1 - \alpha / 2}$
increase $\alpha$ and repeat the previous steps until a new value of
$\alpha$ is obtained so that $\hat{T}_N^{\mbox{mod}} = v_{1 - \alpha^*
/ 2}$, $\alpha^*$ being the p-value of the test.


# Simulations

@SHI:15 provides an example of simulations of non-nested linear models
that shows that the distribution of the Vuong statistic can be very
different from a standard normal. The data generating process used for
the simulations is:

$$
y = 1 + \sum_{k = 1} ^ {K_f} z^f_k + \sum_{k = 1} ^ {K_g} z^g_k + \epsilon
$$

where $z^f$ is the set of $K_f$ covariates that are used in the first
model and $z^g$ the set of $K_g$ covariates used in the second model
and $\epsilon \sim N(0, 1 - a ^ 2)$. $z^f_k \sim N(0, a / \sqrt{K_f})$
and $z^g_k \sim N(0, a / \sqrt{K_g})$, so that the explained variance
explained by the two competing models is the same (equal to $a ^ 2$)
and the null hypothesis of the Vuong test is true. The `vuong_sim`
unables to simulate values of the Vuong test. As in @SHI:15, we use a
very different degree of parametrization for the two models, with $K_f
= 15$ and $K_G = 1$.


```{r }
library(micsr)
Vuong <- vuong_sim(N = 100, R = 1000, Kf = 15, Kg = 1, a = 0.5)
head(Vuong)
mean(Vuong)
mean(abs(Vuong) > 1.96)
```
We can see that the the mean of the statistic for the 1000 replications is
far away from 0, which means that the numerator of the Vuong statistic
is seriously biased. `r round(mean(abs(Vuong) > 1.96) * 100, 1)`% of the
values of the statistic are greater than the critical value so that
the Vuong test will lead in such context a noticeable
overrejection. The empirical pdf is shown in the following figure,
along with the normal pdf.


```{r }
library("ggplot2")
ggplot(data = data.frame(Vuong = Vuong)) + geom_density(aes(x = Vuong)) +
    geom_function(fun = dnorm, linetype = "dotted")
```

# Implementation of the non-degenarate Vuong test

The `micsr` package provides a `ndvuong` function that implements
the classical Vuong test. It has a `nest` argument (that is `FALSE` by
default but can be set to `TRUE` to get the nested version of the
Vuong test). This package also provide a `llcont` generic which
returns a vector of length $N$ containing the contribution of every
observation to the log-likelihood.

The `ndvuong` package provides the `ndvuong` function. As for the
`vuongtest` function, the two main arguments are two fitted models
(say `model1` and `model2`). The $\hat{\Lambda}_n$ vector is obtained
using `llcont(model1) - llcont(model2)`. The relevant matrices $A_i$
and $B_i$ are computed from the fitted models using the `estfun` and
the `meat` functions from the sandwich package. More precisely,
$A^{-1}$ is `bdiag(-bread(model1), bread(model2)`^[`bdiag` if a
function that construct a block-diagonal matrix from its arguments.]
and $B$ is `crossprod(estfun(model1), - estfun(model2)) / N`, where
`N` is the sample size.  Therefore, the `ndvuong` function can be used
with any models for which a `llcont`, a `estfun` and a `bread` method
is available.

# Applications

## Voter turnout

The first application is the example used in @SHI:15 and is used to
compare our **R** program with Shi's **stata**'s
program. @COAT:CONL:04 used several models of electoral participation,
using data concerning referenda about alcool sales regulation in
Texas. Three models are estimated: the prefered group-utilitarian
model, a "simple, but plausible, alternative: the intensity model" and
a reduced form model estimated by the seemingly unrelated residuals
method. They are provided in the `ndvuong` package as `turnout`, a
list of three fitted models^[The estimation is rather complicated
because some linear constraints are used to compute the maximum
likelihood estimator in @COAT:CONL:04's **stata** script. This is the
reason why we provide only the results of the estimations, performed
using the `maxLik` package.]. The results of the Shi test are given
below. We first compute the Shi statistic for an error level of 5%. We
therefore set the `size` argument to 0.05 (this is actually the default
value) and the `pval` argument to FALSE.

```{r }
library("micsr")
test <- ndvuong(turnout$group, turnout$intens, size = 0.05, pval = FALSE)
test
```
The Shi statistic is `r round(unname(test$statistic), 3)`, which is
smaller that the critical value 
`r round(unname(test$parameters["crit-value"]), 3)`. Therefore, based on the Shi test, we
can't reject the hypothesis that the two competing models are equivalent
at the 5% level. The value of the constant $c$ is also reported, as
is the sum of the eigen values of the $V$ matrix (`sum e.v.`). 
The classical Vuong statistic is also reported 
(`r round(unname(test$parameters["vuong_stat"]), 3)`)
and is greater than the 5% normal critical value (the p-value is
`r round(unname(test$parameters["vuong_p.value"]), 3)`).
Therefore, the classical Vuong test and the non-degenerate version lead
to opposite conclusions at the 5% level.

To get only the classical Vuong test, the `nd` argument can be set to
`FALSE`:

```{r }
ndvuong(turnout$group, turnout$intens, nd = FALSE)
```

To get the p-value of the non-degenerate Vuong test, the `pval`
argument should be set to `TRUE`.

```{r }
test <- ndvuong(turnout$group, turnout$intens, pval = TRUE)
test
```
The results indicate that the p-value is `r round(test$p.value, 3)`,
which confirms that the Shi test concludes that the two model are equivalent
at the 5% level.

<!-- ## The demand for medical care (stricly non-nested models) -->

<!-- The second example compares different count models, using data on -->
<!-- demand for medical care. These data were used by @DEB:TRIV:97 and by -->
<!-- @ZEIL:KLEI:JACK:08. They are available as `AER::NMES1988`. The -->
<!-- response is the number of physician office visits (`visits`) and the covariates -->
<!-- are the number of hospital stays (`hosp`), a factor indicating -->
<!-- self-perceived health status (`health`), the number of chronic -->
<!-- conditions (`chronic`), the gender (`gender`), the number of years of -->
<!-- education (`school`) and a factor indicating whether the individual is -->
<!-- covered by private insurance (`insurance`).  -->

<!-- The poisson model is often two simple as the poisson -->
<!-- distribution hypothesis may be rejected for two reasons: -->

<!-- - the first one is that it imposes the equality of the conditional -->
<!--   expectation and variance and that, with real data, we often observe -->
<!--   that the variance is much larger than the expectation, -->
<!-- - the second one is that the probability of 0 for real data are often -->
<!--   much larger that the one implied by the poisson distribution.  -->

<!-- We'll deal here only with the second problem. Two competing models can -->
<!-- be used: -->

<!-- - the zero-inflated (or zip) model, as it names suggests, inflates the -->
<!--   probability of zero, compared to the Poisson model, -->
<!-- - the hurdle model is a two-part model, with a binomial model which -->
<!--   explains the probability of 0 and a zero-truncated poisson model -->
<!--   that explains the probabilities of positive values. -->
  
<!-- These two models can be estimated using respectively `pscl::zeroinfl` -->
<!-- and `pscl::hurdle`. -->

<!-- ```{r } -->
<!-- data("NMES1988", package = "AER") -->
<!-- zi <- pscl::zeroinfl(visits ~ hospital + health + chronic + gender + school + insurance, -->
<!--                      data = NMES1988) -->
<!-- hl <- pscl::hurdle(visits ~ hospital + health + chronic + gender + school + insurance, -->
<!--                    data = NMES1988) -->
<!-- ``` -->

<!-- We first use two previous implementation of the classical Vuong -->
<!-- test. The first one is `pscl::vuong`. -->

<!-- ```{r } -->
<!-- pscl::vuong(zi, hl) -->
<!-- ``` -->

<!-- The alternative hypothesis is writen as the first model being better -->
<!-- than the second one, the p-value is therefore one-sided. The null -->
<!-- hypothesis is therefore rejected at the 5% level (p-value = 0.0143), -->
<!-- and it's also the case if the alternative were writen as one model -->
<!-- being better than the other one (in this case, the p-value would be -->
<!-- $0.0143 \times 2 = 0.0286$). -->

<!-- An other implementation of the classical Vuong test is -->
<!-- `nonnest2::vuong`. -->

<!-- ```{r } -->
<!-- library(nonnest2) -->
<!-- nonnest2::vuongtest(zi, hl) -->
<!-- ``` -->

<!-- This implementation of the test proceeds in two steps. The first step -->
<!-- is the variance test, which is normaly only relevant when the two -->
<!-- models are overlapping. The hypothesis that $\omega ^ 2 = 0$ is -->
<!-- rejected at the 5% level (p-value of 0.0124), so that the second step -->
<!-- is performed. Two alternatives hypothesis are used, the first one -->
<!-- being that the first model is better than the second model and the -->
<!-- second one being the opposite. Therefore, one sided p-values are used -->
<!-- and we get the same conclusion than previously that the zero-inflated -->
<!-- Poisson model is better that the hurdle model at the 5% level. -->

<!-- We then compute the non-degenerate Vuong test: -->

<!-- ```{r } -->
<!-- ndvuong(zi, hl) -->
<!-- ``` -->

<!-- The null hypothesis that the two models are equivalent is not -->
<!-- rejected using the non-degenerate Vuong test, as the p-value is now -->
<!-- equal to 0.057. -->
  
## Transport mode choice (nested models)

The third example concerns transport mode choice in Canada. The
dataset, provided by the `mlogit` package is called `ModeCanada` and
has been used extensively in the transport demand litterature
[see in particular @BHAT:95; @KOPP:WEN:00; and @WEN:KOPP:01]. The
following example is from @CROI:20. The raw data set is first
transformed to make it suitable for the estimation of discrete choice
models. The sample is restricted to the individuals for which 4
transport modes are available (bus, air, train and car). 

```{r }
library("mlogit")
data("ModeCanada")
MC <- mlogit.data(ModeCanada, subset = noalt == 4, chid.var = "case",
                  alt.var = "alt", drop.index = TRUE)
```

We first estimate the simplest discrete choice model, which is the
multinomial logit model. The bus share being negligible, the choice
set is restricted to the three other modes and the reference mode is
set to `car`.

```{r }
ml <- mlogit(choice ~ freq + cost + ivt + ovt | urban + income, MC, 
             reflevel = 'car', alt.subset = c("car", "train", "air"))
```
This model relies on the hypothesis that the unobserved component  of the 
utility functions for the different modes are independent and
identical Gumbell variables. @BHAT:95 proposed the heteroscedastic logit
for which the errors follow a general Gumbell distributions with a
supplementary scale parameter to be estimated. As the overall scale of
utility is not identified, the scale parameter of the reference
alternative (car) is set to one. 


```{r }
hl <- mlogit(choice ~ freq + cost + ivt + ovt | urban + income, MC, 
             reflevel = 'car', alt.subset = c("car", "train", "air"),
             heterosc = TRUE)
coef(summary(hl))
```
The two supplementary coefficients are `sp.train` and `sp.air`. The
student statistics reported are irrelevant because they test the hypothesis
that these parameters are 0, as the relevant hypothesis of
homoscedasticity is that both of them equal one. The heteroscedastic
logit being nested in the multinomial logit model, we can first use
the three classical tests: the Wald test (based on the unconstrained
model `hl`), the score test (based on the constrained model `ml`) and
the likelihood ratio model (based on the comparison of both models). 

To perform the Wald test, we use `lmtest::waldtest`, for which a
special method is provided by the **mlogit** package. The arguments
are the unconstrained model (`hl`) and the update that should be used
in order to get the constrained model (`heterosc = FALSE`). To compute
the scoretest, we use `mlogit::scoretest`, for which the arguments are
the constrained model (`ml`) and the update that should be used in
order to get the unconstrained model (`heterosc = TRUE`). Finally, the
likelihood ratio test is performed using `lmtest::lrtest`.

```{r }
lmtest::waldtest(hl, heterosc = FALSE)
scoretest(ml, heterosc = TRUE)
lmtest::lrtest(hl, ml)
```

The three statistics are $\chi ^2$ with two degrees of freedom under
the null hypothesis of homoscedascticity. The three tests reject the
null hypothesis at the 5% level, and even at the 1% level for the Wald
and the score test.  These three tests relies on the hypothesis that,
under the null, the constrained model is the true model. We can get
rid of this hypothesis using a Vuong test. Note the use of the
`nested` argument that is set to `TRUE`:

```{r }
ndvuong(hl, ml, nested = TRUE)
```

The homoscedasticity hypothesis is still rejected at the 5% level for
the classical Vuong test (the p-value is 0.048), but it is not using
the non-degenerate Vuong test (p-value of 0.211).

## Transport mode choice (overlapping models)

We consider finally another dataset from **mlogit** called
`RiskyTransport`, that has been used by @LEON:MIGU:17 and concerns the
choice of one mode (among water-taxi, ferry, hovercraft and
helicopter) for trips from Sierra Leone's international airport to
downtown Freetown.

```{r }
library("mlogit")
data("RiskyTransport")
RT <- mlogit.data(RiskyTransport, shape = "long", choice = "choice",
                  chid.var = "chid", alt.var = "mode", id.var = "id")
```

We estimate models with only one covariate, the generalized cost of
the mode. We estimate three models: the basic multinomial logit model,
the heteroscedastic model and a nested model where alternatives are
grouped in two nests according to the fact that they are fast or slow
modes.

```{r }
ml <- mlogit(choice ~ cost, data = RT)
hl <- mlogit(choice ~ cost , data = RT, heterosc = TRUE)
nl <- mlogit(formula = choice ~ cost, data = RT,
             nests = list(fast = c("Helicopter", "Hovercraft"),
                          slow = c("WaterTaxi", "Ferry")),
             un.nest.el = TRUE)
modelsummary::msummary(list(multinomial = ml, heteroscedastic = hl, nested =  nl))
```

Compared to ne multinomial model, the heteroscedastic model has 3
supplementary coefficients (the scale parameters for 3 modes, the one
for the reference mode being set to 1) and the nested logit model has
one supplementary parameter which is the nest elasticity (`iv` in the
table). Both models reduce to the multinomial logit model if:

- `sp.WaterTaxi` = `sp.Ferry` = `sp.Hovercraft` = 1 for the
  heteroscedastic model,
- `iv` = 1 for the nested logit model. 

Therefore, the two models are over-lapping, as they reduce to the same
model (the multinomial logit model) for some values of the
parameters. 

The first step of the test is the variance test. It can be performed
using `ndvuong` by setting the argument `vartest` to `TRUE`:

```{r }
ndvuong(nl, hl, vartest = TRUE)
```
The null hypothesis that $\omega^2 = 0$ is rejected. We can then
proceed to the second step, which is the test for non-nested models.

```{r }
ndvuong(hl, nl)
```
The classical and the non-degenerate Vuong tests both conclude that
the two models are equivalent at the 5% level, but that the
heteroscedastic model is better than the nested logit model at the 10% level.



<!-- #' # A poisson model example from the nonnest2 man page -->
<!-- #' data("housing", package = "MASS") -->
<!-- #' house1 <- glm(Freq ~ Infl + Type + Cont, family = poisson, data = housing) -->
<!-- #' house2 <- glm(Freq ~ Infl + Sat,         family = poisson, data = housing) -->
<!-- #' nonnest2::vuongtest(house1, house2) -->
<!-- #' ndvuong(house1, house2) -->
<!-- #' data("bioChemists", package = "pscl") -->
<!-- #' bio1 <- glm(art ~ fem + mar + phd + ment, family=poisson, data=bioChemists) -->
<!-- #' bio2 <- pscl::hurdle(art ~ fem + mar + phd + ment, data=bioChemists) -->
<!-- #' bio3 <- pscl::zeroinfl(art ~ fem + mar + phd + ment, data=bioChemists) -->
<!-- #' nonnest2::vuongtest(bio3, bio2) -->
<!-- #' ndvuong(bio3, bio2) -->

# References {-}

